---
title: "COG Metadata Visualization - Dataset Version 2"
author: "Jenea Adams"
date: "NOV 2022"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---



```{r}
library(dplyr)
library(readr)
library(tidyverse)
# library(mmtable2)
library(ggplot2)

```


# Read in the data 
```{r}
setwd("~/OneDrive - Children's Hospital of Philadelphia/COG-pediatric-AML/metadata_v2")

pat.met = read.csv("patient-met.csv", header = T)
wgs = read.csv("WGS.csv", header = T)
rna = read.csv("RNA.csv", header = T)
# regno = read.csv("regno-to-caseID.csv", header = T)
```


```{r}
setwd("~/OneDrive - Children's Hospital of Philadelphia/COG-pediatric-AML/metadata_v2")
new_pat.met = read.csv("NCT01371981-D4-Dataset.csv")
```

#Inspect the imported data
```{r}
head(new_pat.met)

```
```{r}
head(rna)
```

Change the name of rna case_id --> usi 
```{r}
colnames(rna)[1] = "usi"
head(rna)
```
same for wgs
```{r}
colnames(wgs)[1] = "usi"
head(wgs)

```


#Parse subsets and intersections of the datasets

RNAseq and matched WGS --> 390 
```{r}
length(intersect(rna$usi, wgs$usi))
```

Same as above but using the filter function from dplyr
```{r}
x = filter(rna, rna$usi %in% wgs$usi)

nrow(x)
```



RNA-seq data and patient metadata --> 1040
```{r}

length(intersect(rna$usi, new_pat.met$usi))

```


Sanity check: How many tumor WGS do we have? --> 0 
```{r}
wgs.tum.pat.met = filter(wgs, wgs$usi %in% new_pat.met$usi & !wgs$sample_type == "Normal")
nrow(wgs.tum.pat.met)

```



N individuals with WGS (normal) x RNA-seq tumor x Pat.Met --> 390
```{r}
rna_norm = filter(rna, sample_type == "Normal")

rna_tum = filter(rna, sample_type == "Tumor")

# x2 = filter(rna_tum, rna_tum$usi %in% wgs$usi & rna_tum$usi %in% regno$usi & !rna_tum %in% rna_norm)

rna.wgs = merge(rna, wgs, by="usi")
rna_tum.wgs = merge(rna_tum, wgs, by="usi")
```


```{r}

x2 = filter(new_pat.met, new_pat.met$usi %in% rna_tum$usi & new_pat.met$usi %in% wgs$usi)

nrow(x2)

```
It seems to be that the only intersections of RNA x WGS and Pat.Met are limited by the amount of available WGS data 




How many patients have metadata and EITHER WGS or RNA data? --> 1058 
-- We will use this lateer for downstream visualizations
```{r}

x3 = filter(new_pat.met, new_pat.met$usi %in% rna$usi | new_pat.met$usi %in% wgs$usi)
nrow(x3)

```

How many individuals have patient metadata AND WGS? 
```{r}
x4 = filter(new_pat.met, new_pat.met$usi %in% wgs$usi)
nrow(x4)
```

How many individuals have RNA-seq tumor AND patient metadata 
```{r}
x5 = filter(new_pat.met, new_pat.met$usi %in% rna_tum$usi)
nrow(x5)
```


How many patients have WGS (normal) data but NO RNA-seq or Pat.Met? 
```{r}
x6 = filter(wgs, !wgs$usi %in% rna$usi & !wgs$usi %in% new_pat.met$usi)
nrow(x6)
```


How many patients have RNA-seq tumor but NO RNA-sew tumor or WGS or Pat.Met? 

```{r}
x7 = filter(rna_tum, !rna_tum$usi %in% wgs$usi & !rna_tum$usi %in% new_pat.met$usi & !rna_tum$usi %in% rna_norm$usi)
nrow(x7)
```

How many patients have WGS and RNA-seq tumor only?
```{r}
x8 = filter(rna_tum, rna_tum$usi %in% wgs$usi & !rna_tum$usi %in% new_pat.met$usi & !rna_tum$usi %in% rna_norm$usi)
nrow(x8)
```

How many individuals do we have RNA-seq and pat.met for? 
```{r}
x9 = filter(rna, rna$usi %in% new_pat.met$usi)
nrow(x9)
```

why is this the same number as row 2 in the table? are all of these tumor data? --> yes
```{r}
table(x9$sample_type)
```


# Visualizing patient demographic data 
We will use x3 to represent patients with metadata and some kind of cavatica data (RNA or WGS)
```{r}
cavatica.aaml2 = x3

```


We will look at Race, AgeYr, Risk Group, CBF Identification, years to event free survival, and in the future  WBCl_i for leukemic burden. There are 130+ features by which to model variation across the cohort.

Each of the these is now represented by number indicator variables, and will need to be switched to strings for visualization. 

First let's just define the categories. 
```{r}
race = cavatica.aaml2$race_cat

age = cavatica.aaml2$ageyr

risk = cavatica.aaml2$riskgrp

cbf = cavatica.aaml2$cbf_pt

yrsefs = cavatica.aaml2$yrsefs
```


## Race 
```{r}
race.table = data.frame(table(race))
pie(race.table$Freq, labels = c("AIAN", "Asian", "NHPI", "Black", "White", "MR", "Unknown"), main="Pie chart of updated race categories")

```
```{r}
colnames(race.table)[1] = "Race.Category"
colnames(race.table)[2] = "Count"

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

race.table$fraction = round(race.table$Count/sum(race.table$Count), 3)

race.table$ymax = cumsum(race.table$fraction)

race.table$ymin = c(0, head(race.table$ymax, n=-1))

race_donut_plot = ggplot(race.table, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=Race.Category)) +
  geom_rect() +
  geom_label(aes(label = Count, x = 3.5, y = (ymin+ymax)/2), inherit.aes = T, show.legend = F) +
  coord_polar(theta="y") + # Try to remove that to understand how the chart is built initially
  xlim(c(0, 4)) + theme_void() + scale_fill_discrete(labels = c("AIAN", "Asian", "NHPI", "Black", "White", "MR", "Unknown"))



race_donut_plot

```
Of the known race categories, the proporation of white is 
```{r}
681 / (7+43+5+116+1+681)

```

non-white is 
```{r}
1-0.7983587
```




## Survival Time By Race 
```{r}
library(survival)
library(survminer)
# library(ggplot2)



# fit0 = survfit(Surv(enroll.death.censored, Death.Status) ~1,
#                type = "kaplan-meier",
#                data = cavatica.aaml2)
# summary(fit0)

fit_race = survfit(Surv(yrsos, osi) ~ race_cat,
               type = "kaplan-meier",
               data = cavatica.aaml2)
summary(fit_race)

ggsurvplot(fit_race, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("AIAN", "Asian", "NHPI", "Black", "White", "MR", "Unknown"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           title = "OS time by race", subtitle = "Trial enrollment time to death or censorship")
```
Statistical test of significance between these groups 
```{r}
survdiff(Surv(yrsos, osi) ~ race_cat,
      data = cavatica.aaml2)
```


```{r}
coxph(Surv(yrsos, osi) ~ race_cat,
      data = cavatica.aaml2)
```


```{r}
summary(coxph(Surv(yrsos, osi) ~ race_cat,
      data = cavatica.aaml2))
```

Black vs non-black OS survival 

```{r}

fit_black.nb = survfit(Surv(yrsos, osi) ~ raceb,
               type = "kaplan-meier",
               data = cavatica.aaml2)
summary(fit_black.nb)

ggsurvplot(fit_black.nb, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("Non-Black", "Black"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           palette = c("#52854C", "#D95F02"),
           conf.int = T, 
           title = "OS time by Black vs Non-Black patients", subtitle = "Trial enrollment time to death or censorship")

```


```{r}
summary(coxph(Surv(yrsos, osi) ~ raceb,
               data = cavatica.aaml2))
```


```{r}
fit_black.nb_efs = survfit(Surv(yrsefs, efsi) ~ raceb,
               type = "kaplan-meier",
               data = cavatica.aaml2)
summary(fit_black.nb_efs)

ggsurvplot(fit_black.nb_efs, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("Non-Black", "Black"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           palette = c("#52854C", "#D95F02"),
           conf.int = T, 
           title = "EFS time by Black vs Non-Black patients", subtitle = "Trial enrollment time to death or censorship")
```

```{r}
summary(coxph(Surv(yrsefs, efsi) ~ raceb,
               data = cavatica.aaml2))
```


EFS survival race 
```{r}
fit_race_efs = survfit(Surv(yrsefs, efsi) ~ race_cat,
               type = "kaplan-meier",
               data = cavatica.aaml2)
summary(fit_race_efs)

ggsurvplot(fit_race_efs, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("AIAN", "Asian", "NHPI", "Black", "White", "MR", "Unknown"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           title = "EFS time by race", subtitle = "Trial enrollment time to death or censorship")
```
```{r}
summary(coxph(Surv(yrsefs, efsi) ~ race_cat,
               data = cavatica.aaml2))
```


## Age 
```{r}
hist(age, breaks = 100, main = "Age at Diagnosis", col = heat.colors(50), xlim = c(0,30), 
     xlab = "Age in years", ylab = "Patient frequency")
```


### Survival time by age category
```{r}
fit_age = survfit(Surv(yrsos, osi) ~ agecateg_,
               type = "kaplan-meier",
               data = cavatica.aaml2)
summary(fit_age)

ggsurvplot(fit_age, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("0-1 yo", "2-10 yo", "11-15 yo", "16-20 yo", ">=21 yo"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           title = "OS time by age category", subtitle = "Trial enrollment time to death or censorship")

```
```{r}
summary(coxph(Surv(yrsos, osi) ~ agecateg_,
               data = cavatica.aaml2))
```


## Risk 

```{r}
risk.table = data.frame(table(risk))

```
```{r}
colnames(risk.table)[1] = "Risk.Group"
colnames(risk.table)[2] = "Count"

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

risk.table$fraction = round(risk.table$Count/sum(risk.table$Count), 3)

risk.table$ymax = cumsum(risk.table$fraction)

risk.table$ymin = c(0, head(risk.table$ymax, n=-1))

risk_donut_plot = ggplot(risk.table, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=Risk.Group)) +
  geom_rect() +
  geom_label(aes(label = Count, x = 3.5, y = (ymin+ymax)/2), inherit.aes = T, show.legend = F) +
  coord_polar(theta="y") + # Try to remove that to understand how the chart is built initially
  xlim(c(0, 4)) + theme_void() + scale_fill_discrete(labels = c("Low Risk", "High Risk"))



risk_donut_plot
```

Creating a mosaic plot with ggplot across Race and Risk Group 
```{r}
# # install.packages("devtools")
# # devtools::install_github("haleyjeppson/ggmosaic")
# library(ggmosaic)
```

```{r}
# ggplot(data = cavatica.aaml2) + geom_mosaic(aes(x = product(riskgrp), fill = race_cat)) + theme_mosaic()
```

```{r}
race.risk = table(risk, race)
race.risk
```


```{r}
race.risk_mosaic = mosaicplot(race.risk, xlab = "Risk Classification", ylab = "Race Category", 
                              main = "Risk group classification by race", 
                              col = colorspace::qualitative_hcl(7), 
                              clegend = T, las = 1)

```


### Survival time by risk group

```{r}
fit_risk = survfit(Surv(yrsos, osi) ~ riskgrp,
               type = "kaplan-meier",
               data = cavatica.aaml2)
summary(fit_risk)

ggsurvplot(fit_risk, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("Low Risk", "High Risk"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           conf.int = T,
           palette = c("coral1", "cyan3" ),
           title = "OS time by risk group", subtitle = "Trial enrollment time to death or censorship")
```


Statistical test of significance between the two curves
```{r}
survdiff(Surv(yrsos, osi) ~ riskgrp,
               data = cavatica.aaml2)
```

```{r}
coxph(Surv(yrsos, osi) ~ riskgrp,
               data = cavatica.aaml2)
```


```{r}
cph.fit.risk = coxph(Surv(yrsos, osi) ~ riskgrp,
               data = cavatica.aaml2)
summary(cph.fit.risk)
```

```{r}

temp = cox.zph(cph.fit.risk)
temp

plot(temp)
```


Looking at EFS --> are there more types of events in patients across rislk groups? 
```{r}
fit_risk_efs = survfit(Surv(yrsefs, efsi) ~ riskgrp,
               type = "kaplan-meier",
               data = cavatica.aaml2)
summary(fit_risk_efs)

ggsurvplot(fit_risk_efs, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("Low Risk", "High Risk"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           conf.int = T,
           palette = c("cyan3", "coral1"),
           title = "EFS time by risk group", subtitle = "Trial enrollment time to death or censorship")
```

```{r}
summary(coxph(Surv(yrsos, osi) ~ riskgrp,
               data = cavatica.aaml2))
```


## CBF
```{r}
table(cbf)
```
```{r}
race.cbf = table(cbf, race)

race..cbf_mosaic = mosaicplot(race.cbf, xlab = "CBF Mutation", ylab = "Race Category", 
                              main = "CBF Mutation by Race", 
                              col = colorspace::qualitative_hcl(7), 
                              clegend = T, las = 1)
```


Filtering the matients that have a CBF mutation --> 244 patients 
```{r}
cbf.mut.pts = filter(cavatica.aaml2, cbf =="CBF")

cbf.race.table = data.frame(table(cbf.mut.pts$race_cat))

colnames(cbf.race.table)[1] = "Race.Category"
colnames(cbf.race.table)[2] = "Count"



cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

cbf.race.table$fraction = round(cbf.race.table$Count/sum(cbf.race.table$Count), 3)

cbf.race.table$ymax = cumsum(cbf.race.table$fraction)

cbf.race.table$ymin = c(0, head(cbf.race.table$ymax, n=-1))

cbf.race_donut_plot = ggplot(cbf.race.table, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=Race.Category)) +
  geom_rect() +
  geom_label(aes(label = Count, x = 3.5, y = (ymin+ymax)/2), inherit.aes = T, show.legend = F) +
  coord_polar(theta="y") + # Try to remove that to understand how the chart is built initially
  xlim(c(0, 4)) + theme_void() + scale_fill_discrete(labels = c("AIAN", "Asian", "NHPI", "Black", "White", "Unknown"))



cbf.race_donut_plot

```


This shows that no MR patients have a CBF mutation. The proportion of white from the known are
```{r}
cbf.race.table
```

```{r}
170 / (1+12+2+29+170)
```


### Survival Time by CBF status

First, we need to create a dummy variable for CBF mutation status in the main dataframe 
```{r}
cavatica.aaml2$cbf_ind = ifelse(cavatica.aaml2$cbf_pt =="CBF", 1, 0)
head(cavatica.aaml2)
```

Now that we have an indicator variable for cbf status, we can create the survival plot against this feature 
```{r}


fit_cbf = survfit(Surv(yrsos, osi) ~ cbf_ind,
               type = "kaplan-meier",
               data = cavatica.aaml2)


summary(fit_cbf)

ggsurvplot(fit_cbf, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("No CBF mutation", "CBF mutation"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           conf.int = T,
           palette = c("cyan3", "coral1"),
           title = "OS time by CBF mutation status", subtitle = "Trial enrollment time to death or censorship")
```


Per findings by Rau et al, CBf mutations seem to predict better survival (I'm looking at OS, but they compare EFS). 

Let's look at EFS to make sure. 

```{r}
library(wesanderson)
```


```{r}
fit_cbf_efs = survfit(Surv(yrsefs, efsi) ~ cbf_ind,
               type = "kaplan-meier",
               data = cavatica.aaml2)


summary(fit_cbf_efs)

ggsurvplot(fit_cbf_efs, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("No CBF mutation", "CBF mutation"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           conf.int = T,
           palette = c("#999999", "#CC79A7"),
           title = "EFS time by CBF mutation status", subtitle = "Trial enrollment time to death or censorship")
```


Performing statistical tests to see if there is a significant difference between the curves 
https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html

- We can conduct between-group significance tests using a log-rank test
- The log-rank test equally weights observations over the entire follow-up time and is the most common way to compare survival times between groups
- There are versions that more heavily weight the early or late follow-up that could be more appropriate depending on the research question (see ?survdiff for different test options)

We get the log-rank p-value using the survdiff function.


```{r}
sd_cbf_efs = survdiff(Surv(yrsefs, efsi) ~ cbf_ind,
                      data = cavatica.aaml2)
sd_cbf_efs
```

extracting the actual p-value 
```{r}
1 - pchisq(sd_cbf_efs$chisq, length(sd_cbf_efs$n) - 1)
```


Cox regression modelling 

We may want to quantify an effect size for a single variable, or include more than one variable into a regression model to account for the effects of multiple variables.

The Cox regression model is a semi-parametric model that can be used to fit univariable and multivariable regression models that have survival outcomes.

h(t|Xi)=h0(t)exp(β1Xi1+⋯+βpXip)

h(t): hazard, or the instantaneous rate at which events occur h0(t): underlying baseline hazard

Some key assumptions of the model:

- non-informative censoring
- proportional hazards

Note: parametric regression models for survival outcomes are also available, but they are not addressed in this training

We can fit regression models for survival data using the coxph function, which takes a Surv object on the left hand side and has standard syntax for regression formulas in R on the right hand side.
```{r}
coxph(Surv(yrsefs, efsi) ~ cbf_ind,
      data = cavatica.aaml2)
```


```{r}
summary(coxph(Surv(yrsefs, efsi) ~ cbf_ind,
      data = cavatica.aaml2))
```

Let's visualize in nicer table?
```{r}
# coxph(Surv(yrsefs, efsi) ~ cbf_ind,
#       data = cavatica.aaml2) %>% gtsummary::tbl_regression(exp = TRUE)

```


```{r}
survdiff(Surv(yrsefs, efsi) ~ cbf_ind,
      data = cavatica.aaml2)
```



Similar to Rau et al, I'd like to take a look patients with CBF mutations to see if survival differs between Black and non-Black patients 

```{r}
fit_cbf.only_efs = survfit(Surv(yrsefs, efsi) ~ raceb,
               type = "kaplan-meier",
               data = cbf.mut.pts)


summary(fit_cbf.only_efs)

ggsurvplot(fit_cbf.only_efs, data = cbf.mut.pts,
           legend="bottom",
           legend.labs = c("Non-Black", "Black"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           conf.int = T,
           palette = c("#52854C", "#D95F02"),
           title = "EFS time in CBF mutant patients by Black or Non-Black race groups", subtitle = "Trial enrollment time to death or censorship")
```

```{r}
nrow(cbf.mut.pts)
```


Statistical test on this subset

```{r}
summary(coxph(Surv(yrsefs, efsi) ~ raceb,
      data = cbf.mut.pts))
```

```{r}
survdiff(Surv(yrsefs, efsi) ~ raceb,
      data = cbf.mut.pts)
```


With this smaller group of patients, it does not appear that there is a significant difference of EFS between Black and Non-Black patients

## NPM
```{r}
table(cavatica.aaml2$npmstat_)
```


```{r}
npm.mut.pts = filter(cavatica.aaml2, cavatica.aaml2$npmstat_ == 1)

fit_npm_efs = survfit(Surv(yrsefs, efsi) ~ npmstat_,
               type = "kaplan-meier",
               data = cavatica.aaml2)


summary(fit_npm_efs)

ggsurvplot(fit_npm_efs, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("NPM mutation", "No NPM mutation"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           conf.int = T,
           palette = c("#00AFBB", "#E7B800"),
           title = "EFS time in NPM mutant patients", subtitle = "Trial enrollment time to death or censorship")


```

```{r}
summary(coxph(Surv(yrsefs, efsi) ~ npmstat_,
               data = cavatica.aaml2))
```


This paper on the 1031 trial (https://chop365-my.sharepoint.com/personal/adamsj8_chop_edu/Documents/COG-pediatric-AML/literature/AML_Pediatric_Sorafenib_AAML1031_Pollard_JCO_2022.pdf) reports that co-ocurring mutations of FLT3 and NPM seem to improve survival. Does NPM+ mutation improve survival of those with FLT mutations? 

```{r}

# cavatica.aaml2$npm.flt = ifelse(cavatica.aaml2$npmstat_ == 1, 1, (ifelse(cavatica.aaml2$itdlowHAR ==)))


# cavatica.aaml2$npm.flt = NA
# for(i in length(cavatica.aaml2)) {
#   if(cavatica.aaml2$npmstat_ == 1 && cavatica.aaml2$itdlowHAR == 2){cavatica.aaml2$npm.flt = 0}
#   else if(cavatica.aaml2$npmstat_ == 2 && cavatica.aaml2$itdlowHAR == 1){cavatica.aaml2$npm.flt = 1}
#   else {cavatica.aaml2$npm.flt = 2}
# }

cavatica.aaml2 %>% mutate(npm.flt = case_when(
  npmstat_ == 1 && itdlowHAR == 2 ~ 0, 
  npmstat_ == 2 && itdlowHAR == 1 ~ 1,
  npmstat_ == 1 && itdlowHAR == 1 ~ 2,
  npmstat_ == 2 && itdlowHAR == 2 ~ 0,
))


head(cavatica.aaml2$npm.flt)

```



This paper on the 1031 trial (https://chop365-my.sharepoint.com/personal/adamsj8_chop_edu/Documents/COG-pediatric-AML/literature/AML_Pediatric_Sorafenib_AAML1031_Pollard_JCO_2022.pdf) reports NPM mutations being associated with higher WBC, is this true? --> not exactly? 

```{r}
# library(vioplot)


boxplot(data = cavatica.aaml2, log(wbc1)~npmstat_, 
        names = c("NPM +", "NPM -"))
```

```{r}
cavatica.aaml2$npm = ifelse(cavatica.aaml2$npmstat_ == 1, "NPM+", "NPM-")

ggplot(cavatica.aaml2 %>% filter(!is.na(npm)), aes(x=npm, y=log(wbc1))) + geom_violin(aes(fill=npm)) + 
  geom_boxplot(aes(alpha = 0.5), width=.4) + 
  geom_jitter(width = 0.3) + 
  scale_fill_manual(values = c("#00AFBB", "#E7B800")) + 
  ggtitle("Log(WBC) across NPM mutation groups", subtitle = "N=956 complete observations")
  
  
# ggplot(cavatica.aaml2, aes(x=npm, y=wbc1)) + geom_dotplot(binaxis='y', stackdir='center', dotsize = 1)
```
```{r}
ggplot(cavatica.aaml2 %>% filter(!is.na(npm)), aes(x=npm, y=log(wbc1))) + geom_violin(aes(fill=npm)) + 
  geom_boxplot(aes(alpha = 0.5), width=.4) + 
  geom_jitter(width = 0.3) + 
  scale_fill_manual(values = c("#00AFBB", "#E7B800")) + 
  ggtitle("Log(WBC) across NPM mutation groups", subtitle = "N=956 complete observations") + 
  stat_compare_means(method = "t.test", label.x = 1.5)
```


```{r}
table(cavatica.aaml2$npm)
```
```{r}
npm_pos = filter(cavatica.aaml2, cavatica.aaml2$npm == "NPM+")
npm_neg = filter(cavatica.aaml2, cavatica.aaml2$npm == "NPM-")
```

```{r}
hist(npm_pos$wbc1)
```


```{r}
mean(npm_pos$wbc1)
```

```{r}
mean(npm_neg$wbc1)
```


```{r}
t.test(npm_neg$wbc1, npm_pos$wbc1, paired = F, var.equal = F, alternative = "less")
```


```{r}
t.test(npm_neg$wbc1, npm_pos$wbc1, paired = F, var.equal = F, alternative = "greater")
```


## Leukemic Burden --> WBC groups 

```{r}
table(cavatica.aaml2$wbc1_i)
```


```{r fit wbc efs}

fit_wbc_efs = survfit(Surv(yrsefs, efsi) ~ wbc1_i,
               type = "kaplan-meier",
               data = cavatica.aaml2)


summary(fit_wbc_efs)

ggsurvplot(fit_wbc_efs, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("<=100,000/μL", ">100,000/μL", "Unknown"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           conf.int = T,
           # palette = c("cyan3", "coral1"),
           title = "EFS time across WBC category", subtitle = "Trial enrollment time to death or censorship")
```


High white blood cell count at presentation is an unfavorable prognostic factor for treatment outcome (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3166101/). In this case, this higher burden seems to result in decrease median survival between these groups. 

```{r}
survdiff(Surv(yrsefs, efsi) ~ wbc1_i,
               data = cavatica.aaml2)
```

This seems to be supported by a significant p-value between the groups. 


Leukemic burden across racial groups 
```{r}

cavatica.aaml2 = cavatica.aaml2 %>% mutate(race_str = case_when(
  race_cat == 1 ~ "AIAN", 
  race_cat == 2 ~ "Asian",
  race_cat == 3 ~ "NHPI",
  race_cat == 4 ~ "Black",
  race_cat == 5 ~ "White",
  race_cat == 6 ~ "MR", 
  race_cat == 9 ~ "Unknown"
))

ggplot(cavatica.aaml2 %>% filter(!is.na(race_str)), aes(x=race_str, y=log(wbc1))) + 
  geom_violin(aes(fill =race_str)) + 
  geom_boxplot(aes(alpha = 0.4), width=.35) +
  geom_jitter(width = 0.3, aes(alpha = 0.4)) + 
  # scale_fill_manual(values = c("#00AFBB", "#E7B800")) + 
  ggtitle("Log(WBC) across Race categories", subtitle = "N=956 complete observations") + 
  stat_compare_means(method = "kruskal.test", label.x = 1.5) + 
  guides(alpha = F) + 
  scale_fill_discrete(name = "Race") + 
  xlab("Race") + 
  ylab("log(WBC)")

```

I decided to stick with the Kruskal-Wallis test because it's non-parametrics and does not assume homoscedasticity within the data. It assumes that the groups come from the same distribution. 

## Gender 

```{r}
fit_gender_os = survfit(Surv(yrsos, osi) ~ gender,
               type = "kaplan-meier",
               data = cavatica.aaml2)


summary(fit_gender_os)

ggsurvplot(fit_gender_os, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("Male", "Female"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           conf.int = T,
           # palette = c("cyan3", "coral1"),
           title = "OS time across Gender", subtitle = "Trial enrollment time to death or censorship")
```

```{r}
summary(coxph(Surv(yrsos, osi) ~ gender,
               data = cavatica.aaml2))
```

There is not enough evidnece to support a significant difference in overall survival across gender groups. 



## Treatment Arm 

```{r}
fit_trtArm_os = survfit(Surv(yrsos, osi) ~ trt_arm,
               type = "kaplan-meier",
               data = cavatica.aaml2)


summary(fit_trtArm_os)

ggsurvplot(fit_trtArm_os, data = cavatica.aaml2,
           legend="bottom",
           legend.labs = c("Arm A", "Arm B"),
           ggtheme = theme_bw(),
           xlab = "Time (Years)",
           conf.int = T,
           # palette = c("cyan3", "coral1"),
           title = "OS time across treatment arms", subtitle = "Trial enrollment time to death or censorship")
```

```{r}
summary(coxph(Surv(yrsos, osi) ~ trt_arm,
               data = cavatica.aaml2))
```



# Multivariate cox regression on survival times across features 
```{r}
res.cox1 = coxph(Surv(yrsefs, efsi) ~ race_cat + riskgrp, 
                 data = cavatica.aaml2)
summary(res.cox1)
```


```{r}

res.cox2 = coxph(Surv(yrsefs, efsi) ~ as.character(race_cat) + as.character(riskgrp) + as.character(npmstat_) + cbf_pt + as.character(itdlowHAR), 
                 data = cavatica.aaml2)
summary(res.cox2)
```


visualizing these reuslts: https://rpkgs.datanovia.com/survminer/survminer_cheatsheet.pdf 

https://shariq-mohammed.github.io/files/cbsa2019/1-intro-to-survival.html#6_cox_regression 


```{r}
res.cox2 = coxph(Surv(yrsefs, efsi) ~ as.character(race_cat) + as.character(riskgrp) + as.character(npmstat_) + cbf_pt + as.character(itdlowHAR), 
                 data = cavatica.aaml2)
summary(res.cox2)
```

Testing the proportional hazards assumption : In principle, the Schoenfeld residuals are independent of time. A plot that shows a non-random pattern against time is evidence of violation of the PH assumption. 
For each covariate, the function cox.zph() correlates the corresponding set of scaled Schoenfeld residuals with time, to test for independence between residuals and time. Additionally, it performs a global test for the model as a whole. The proportional hazard assumption is supported by a non-significant relationship between residuals and time, and refuted by a significant relationship.

```{r}
test.ph1 = cox.zph(res.cox2)
test.ph1
```

The proportional hazard assumption is not supported for the risk group feature since there is a significant relationship between residuals and time. Proportional hazards assumption for this model is not reasonable (?)


```{r}
ggcoxzph(test.ph1)

```


Looking for outliers 
```{r}
ggcoxdiagnostics(res.cox2, type = "dfbeta", linear.predictions = F, 
                 ggtheme = theme_bw())
```

```{r}
ggcoxdiagnostics(res.cox2, type = "deviance", linear.predictions = F, 
                 ggtheme = theme_bw())
```

It is also possible to check outliers by visualizing the deviance residuals.

The deviance residual is a normalized transform of the martingale residual.

These residuals should be roughly symmetrically distributed about zero with a standard deviation of 1.

Positive values correspond to individuals that “died too soon” compared to expected survival times.

Negative values correspond to individual that “lived too long”.

Very large or small values are outliers, which are poorly predicted by the model.

```{r}
# res.cox3 = coxph(Surv(yrsefs, efsi) ~ race_cat + riskgrp + npmstat_ + cbf_pt + itdlowHAR, 
#                  data = cavatica.aaml2)
# summary(res.cox3)
```


A forest plot to summarize the results across these features for this particular model. 
How to read a forest plot: https://s4be.cochrane.org/blog/2016/07/11/tutorial-read-forest-plot/ 

```{r}
# ggforest(res.cox2)

```

The bigger the black box, the more participants in this group. 
The horzontal line through the group represents the 95% confidence interval




adjusted survival curves 
```{r}
# ggcoxadjustedcurves(res.cox2, data = cavatica.aaml2)
```


----
# Preparing for rMATS-turbo 
Once exploratory data analysis is complete, Export the list of data subsets needed and group them in Cavatica for downstream analysis (move_riskgroup_bams_cavatica_args.py)

